---
title: 'Using R: The Basics'
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# Variables

## Setting Variables

You’ll start wanting to save data/information as variables. This is done using `<-`

You can see your variables in the Environment Tab of the RStudio Panel.

```{r}
x <- 2
x
y <- x + 3
y
```

## Types of Variables

```{r echo = FALSE, warning = FALSE}
library(png)
library(grid)
img <- readPNG("Rmd/img/R-data-structure.png")
grid.raster(img)
```

## Examples

```{r}
# Numeric vector
x <- c(1, 2, 6, 10)
x
# 2 x 3 String Matrix
m <- c("R", "is", "the", "best", "stats", "program")
m <- matrix(m, nrow = 2)
m
```

## Examples

```{r}
# A Data frame
sex <- c("m", "m", "f", "f")
age <- c(50, 30, 22, 45)
height <- c(170, 160, 180, 174)
d <- data.frame(sex, age, height)
d
```

# Functions

The vocabulary of R

## What are Functions?

Functions encapsulate a series of commands, executed top to bottom, left to right, to perform a task and return a result.

`mean()`, for example, is a function. It adds up all the elements of a vector, then divides them by the number of elements, and returns the result.

## Recognising a Function

Functions all have the same structure when called:

- A function name
- An open parenthesis `(`
- Comma separated arguments/inputs (if any)
- Close parenthesis `)`

Example:

```{r}
v <- c(10, 20, 59.8, 299)  # c() is a function for creating vectors
mean(v)
```

## Base functions

There are lots of other functions already in R.

Try these for yourself:

```{r, eval = F}
v <- c(5, 10, 15,20)
length(v)
sd(v)
sum(v)
range(v)
summary(v)
```

Recall, you can get help by running the command  
`?functionName`

## Custom functions

We can also write our own functions in this format:

```{r, eval = F}
functionName <- function(arg1, arg2, ..., argn) {
  command 1
  command 2
  ...
  command n
  return (result)
}

```

Once it's in our environment, we can call our function like this:

```{r, eval = F}
functionName(arg1, arg2, ..., argn)
```

## Custom Functions

Here's a custom function for `mean()`:

```{r}
cMean <- function(v) {
  s <- sum(v)
  n <- length(v)
  m <- s / n
  return (m)
}

cMean(c(1, 2, 3))
cMean(c(11, 27, 24.32, 89, 102))
```

# Input and Output

## Reading Tabular Data

Tabular data can be read with `read.table("path.to.data")` and variants. Most frequent variant is `read.csv()`

```{r}
d <- read.csv("data/risk.csv")  # risk.csv is course material
```
```{r, eval = FALSE}
View(d)  # Run this yourself
```
```{r}
class(d)
dim(d)
```

## Reading Data

`*.csv` generally easiest, but you can import all sorts of data types:

- Excel, SPSS, SAS, STATA, tab delimited, anything delimited, Fixed width, HTML (web scraping), MySQL, etc

For alternatives, consult other resources such as [Quick-R](http://www.statmethods.net/input/importingdata.html).

## Writing data

You can easily save matrix or data frame variables as `*.csv` using `write.csv()`

Example:

```{r, eval = F} 
m <- matrix(c(1, 2, 3, 4), nrow = 2)
write.csv(m, file = "myMatrix.csv")
```

Of course, other options exist.
