---
title: "Using R: Data Frame tidying with tidyr"
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# tidyr

## What is tidyr?

tidyr is an R package that makes it easy to "tidy" your data frames.

It has a core set of "verbs" (functions) that handle most of our essential needs.

In this session, we'll look at the following:

- `separate()`
- `gather()`
- `spread()`

## Getting tidyr

Start by installing tidyr (if you don't have it already), and loading it into your workspace.

```{r, eval = F}
# Install the package from CRAN (if necessary)
install.packages("tidyr")

# Load it into your workspace
library(tidyr)
```
```{r, include = F}
library(tidyr)
```

## and also...

We'll also be using `dplyr`


```{r, eval = F}
# Install the package from CRAN (if necessary)
install.packages("dplyr")

# Load it into your workspace
library(dplyr)
```
```{r, include = F}
library(dplyr)
```

## Function Style

Just like functions in `dplyr`, the "verbs" of tidyr all share a similar format:

- their first arguement is a data frame.
- subsequent arguments are specific to the function.
- they all return a data frame.

When referencing column names, we don't need quotation marks ""

# Time to tidy...

## Separate a single column with `separate()`

```{r}
d <- "data/flight-times.csv" %>%
  read.csv() %>%
  select(flight, origin, dest)
head(d)
```

## Separate a single column with `separate()`

```{r}
d %>%
  separate(flight, into = c("carrier", "flight_num"), sep = "-") %>%
  head()
```


## Gather multiple columns together with `gather()`

Convert multiple columns into two columns: a key and a value.

Useful for going from wide format to long format.

`gather(data, key.name, value.name, columns.to.gather)`

## Gather multiple columns together with `gather()`

```{r}
d <- read.csv("data/energy.csv")
d %>% gather(exercise, energy, Running, Walking, Cycling)
```

## Gather multiple columns together with `gather()`

If you have adjacent columns, you can make life easier with `:`

```{r}
d <- read.csv("data/energy.csv")
d %>% gather(exercise, energy, Running:Cycling)
```

## Spread key-value columns with `spread()`

Convert two columns, representing a key and value, into multiple columns.

Useful for getting from long to wide format.

## Spread key-value columns with `spread()`

```{r}
d <- "data/risk.csv" %>%
  read.csv %>%
  select(subid, active, risk) %>%
  filter(subid < 20 & active %in% c("GOLD", "SILV"))
d
```

## Spread key-value columns with `spread()`

```{r}
d %>% spread(active, risk)
```