---
title: "Using R: Data Frame M&Ms with dplyr"
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# dplyr

## What is dplyr?

dplyr is an R package that makes data frame M&Ms easy, simple, clean and efficient.

It has a core set of "verbs" (functions) that handle most of our essential needs.

In this session, we'll look at the following:

- `select()`
- `filter()`
- `mutate()`
- `arrange()`
- `join()`
- Pipelines with `%>%`

## Getting dplyr

Start by installing dplyr (if you don't have it already), and loading it into your workspace.

```{r, eval = F}
# Install the package from CRAN (if necessary)
install.packages("dplyr")

# Load it into your workspace
library(dplyr)
```
```{r, include = F}
library(dplyr)
```

## Our Data Set

Let's read in the risk data to practice.

```{r}
d <- read.csv("data/risk.csv")
head(d)
```

# dplyr Basics

## Function Style

The "verbs" of dplyr all share a similar format:

- their first arguement is a data frame.
- subsequent arguments are specific to the function.
- they all return a data frame.

When referencing column names, we don't need quotation marks ""

## How to Select Columns

With dplyr, we can select columns using `select()`

```{r}
x <- select(d, risk)
head(x)
```

## How to Select Columns

Use the dash `-` to remove columns

```{r}
x <- select(d, -risk)
head(x)
```

## How to Select Columns

We can add as many column names as we like.
```{r}
x <- select(d, risk, gender, age)
head(x)
```

## How to Filter Rows

With dplyr, we can select certain rows using `filter()`

```{r}
x <- filter(d, risk > 90)
head(x)
```

## How to Filter Rows

These can be compounds...

```{r}
x <- filter(d, risk > 90 & age == 20)
head(x)
```

## Create New Variables

New variables can be created with `mutate()`

```{r}
x <- mutate(d, risk.sq = risk^2)
head(x)
```

## Create New Variables

We can create, repalce, and remove multiple variables at once!

```{r}
x <- mutate(d, risk.sq = risk^2)
x <- mutate(x, risk.cu = risk^3,
               gender  = factor(gender, levels = c(0, 1),
                                labels = c("female", "male")),
               risk.sq = NULL)
str(x)
```

## Create New Variables

Aside, we can rename columns with `rename()`

```{r}
x <- rename(d, sex = gender, world.view = wvcat)
head(x)
```

## Arrange Rows

We can arrange rows with `arrange()`

```{r}
head(arrange(d, age))
```

## Arrange Rows

We can arrange in descending order with `desc`

```{r}
head(arrange(d, desc(age)))
```

## Arrange Rows

Break ties with more arguments.

```{r}
head(arrange(d, age, desc(risk)))
```

## Combining What We Know

```{r}
x <- mutate(d, risk.sq = risk^2)
x <- select(x, gender, age, risk.sq)
x <- filter(x, age > 80)
x <- arrange(x, age, risk.sq)
head(x)
```

## Join Two Data Sets

We can join data sets with a set of functions of the form `xxx_join(df1, df2, by, ...)`

Similar to `merge()`, but with various advantages.

```{r}
x <- select(d, subid, active, risk)
y <- select(d, subid, active, wvcat)
left_join(x, y)
```

# Pipelines

## What is going on here?

```{r, eval = F}
arrange(
  filter(
    select(
      mutate(d, risk.sq = risk^2),
      gender, age, risk.sq
    ),
    age > 80
  ),
  age, risk.sq
)
```

## Why Pipelines?

Help us to streamline our code and make it readable.

In dplyr (and associated packages), we "pipe" variables with `%>%`

The format is `x %>% f(y, ...)`

...which implicity becomes `f(x, y, ...)`

## Examples

```{r}
d %>% head()
```

## Examples

```{r}
d %>% filter(wvcat != 1)
```

## Examples

```{r}
d %>%
  select(gender, age, risk) %>%
  filter(risk < 5)
```

## Examples

Trick: read it aloud as "then".

```{r}
d %>%
  mutate(risk.sq = risk^2) %>%
  select(gender, age, risk.sq) %>%
  filter(age > 80) %>%
  arrange(age, risk.sq)
```

# Some Base Comparisons

## Selecting Columns

The following are equivalent:

```{r, eval = F}
d[, c("x", "y", "z")]  # Base R

select(d, x, y, z)     # dplyr, no piping

d %>% select(x, y, z)  # dplyr, with piping
```

## Filtering Rows

The following are equivalent:

```{r, eval = F}
d[d$x == "a" & d$y < 100, ]

filter(d, x == "a" & y < 100)

d %>% filter(x == "a" & y < 100)
```

## Filtering Rows

The following are equivalent:

```{r, eval = F}
d[d$x == "a" & d$y < 100, ]

filter(d, x == "a" & y < 100)

d %>% filter(x == "a" & y < 100)
```

## Create New Variables

The following are equivalent:

```{r, eval = F}
d$x2 <- d$x ^ 2

d <- mutate(d, x2 = x^2)

d <- d %>% mutate(x2 = x^2)
```

## Merge/Join data frames

```{r, eval = F}
df1 <- data.frame(id = c(1, 2), age = c(45, 32))
df2 <- data.frame(id = c(1, 2), gender = c("f", "m"))
```

The following are equivalent:

```{r, eval = F}
merge(df1, df2)

left_join(df1, df2)

df1 %>% left_join(df2)
```

## Combinations

The following are equivalent:

```{r, eval = F}
d <- d[d$x == "a" & d$y < 100, c("x", "y", "z")]
d$x2 <- d$x ^ 2

###

d <- select(d, x, y, z)
d <- filter(d, x == "a" & y < 100)
d <- mutate(d, x2 = x^2)
```

Continues on next slide...

## Combinations

```{r, eval = F}

d <- mutate(
      filter(
        select(d, x, y, z)
        , x == "a" & y < 100
        )
      , x2 = x^2
      )

###

d <- d %>%
  select(x, y, z) %>%
  filter(x == "a" & y < 100) %>%
  mutate(x2 = x^2)
```

