---
title: "Using R: Data Frame M&Ms"
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# M&Ms: Management and Manipulations

## Our Data Set

Let's read in the risk data to practice.

```{r}
d <- read.csv("data/risk.csv")
head(d)
```

# What's in our Data Frame?

## Functions for Examining Data Frames

We've seen some basics:

```{r, eval = F}
View(d)
dim(d)
head(d)
```

But how else can we examine information about our data?

## Functions for Examining Data Frames

```{r}
colnames(d)  # Columns names
str(d)  # Display Internal Structure
```

Notice many categorical variables are integers. We should change these to factors (later).

# Subsetting

## Selecting rows and columns

data.frame rows and columns can be accessed with square brackets like `data.frame[rows, columns]`

We can do this using numeric indices (i.e., the i*th* row), character names (e.g., columns "gender"), or boolean values (more later).

Try these for yourself:

```{r, eval = F}
d[1, ]  # The first row
d[, 5]  # The fifth column
d[, "age"]  # The column called "age""
d[1, "age"]  # First row, "age" column
```

You can also use `$` as a short hand for selecting **one column**. E.g., `d$age`

## Selecting rows and columns

Vectors can be used to select multiple rows and/or columns.

```{r}
row.vec <- c(1, 10, 20)  # Rows to select
col.vec <- c("subid", "gender", "age")  # Columns to select
d[row.vec, col.vec]
```

## Conditionals

Selecting rows based on conditions involves two steps (though they can be combined into one line of code).

1. Use a conditional statement to create a boolean vector corresponding to the rows of your data. This vector should be TRUE for the rows you wish to select.
2. Use this vector in to select the rows.

## Conditional Row Selection

Step 1 - creating a boolean vector corresponding to the rows using operators and functions like  
`== != < > <= >= & | is.na() ! %in%`

For example, try:

```{r, eval = F}
d$risk > 80  # All rows where risk is grater than 80
d$subid == 15  # All rows where the subject id is 15
d$risk <= 20 & d$ethnic != 1  # risk less than or equal to 20 AND
                              # ethnicity not equal to 1
d$risk < 20 | d$risk > 90  # risk less than 20 OR greater than 90
```

For convenience, you can assign this to a vector variable:

```{r}
row.vec <- d$subid == 917 & d$risk > 30
```

## Conditional Selection

Step 2 - use the conditional to select your rows.

For example:

```{r}
row.vec <- d$subid == 917 & d$risk > 30
d[row.vec, ]
```

This can also be combined with columns

```{r}
d[row.vec, "risk"]
```

## Test yourself

Subset the data to get all rows with a risk rating of less than 20, OR greater than or equal to 90, but NOT 100.

## The Answer

```{r}
row.vec <- d$risk < 20 | d$risk >= 90 & d$risk != 100
hist(d[row.vec, "risk"])  # Plot a Histogram of risk column
```

# Manipulating Data

## Changing data in R

Sometimes we need to make changes to our data (e.g., change values, create new columns, etc).

One option is to do it manually (e.g., in the .csv file) on in another program.

But this goes against The Golden Rule 1 of Function Science!

We'll look at some common changes.

## Changing a cell

Changing data is like combining selection and variable assignment.

For example, we can assign a value directly to a particular cell.

```{r}
d[1, "risk"] <- 37
d[1, ]
```

## Changing subsets of data

The same principle applies for subsets of data

```{r}
row.vec <- !is.na(d$risk) & d$risk == 100
d[row.vec, "risk"] <- NA
head(d[row.vec,])
```

## Overwrite and Create Columns

Overwrite or create a column by assigning it a value.

```{r}
d$subid <- d$subid + 1  # Overwrite subid as itself plus 1
d$zero <- 0  # a column of all zeros
d$age.100 <- d$age + 100  # Add 100 to everyone's age
d$risk2 <- d$risk^2  # risk squared
head(d)
```

## Converting to factor

Convert to a factor with `factor()`, select your levels and give labels if needed. Be sure to check the "ordering".

```{r}
d$gender <- factor(d$gender,  # A vector (itself in this case)
                   levels = c(0, 1),  # Unique values in vector
                   labels = c("female", "male"))  # Labels to assign
d[c(1, 14, 670), "gender"]
```
