---
title: "Descriptive Statistics"
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# Setup

## Packages

We'll make use of the `psych` package.

Make sure it's installed on you system and then load it into your workspace.

```{r, eval = F}
# Install the package from CRAN (if necessary)
install.packages("dplyr")

# Load it into your workspace
library(dplyr)
```
```{r, include = F}
library(dplyr)
```

## Our Data Set

Let's read in the flight times data to practice.

```{r}
d <- read.csv("data/flight-times.csv")
head(d)
```

# Grouping and Summarising

## Splitting data into groups

Often we're interested in splitting our data into groups.

We do this with the dplyr function, `group_by()`

For example:

```{r}
by.fn <- group_by(d, ORIGIN)
```

Other dplyr functions will now treat `by.fn` as multiple data frames: one for each level of the FL_NUM factor.

More to come...

## Summarising Data

To summarise information about a data frame, we use the dplyr function, `summarise()`

```{r}
summarise(d, mean = mean(CRS_DEP_TIME))
summarise(d, mean = mean(CRS_DEP_TIME),
             max  = max(CRS_DEP_TIME))
```

# Univariate Analysis

## Frequencies and Proportions

For the frequency of a factor variable, group it, and then summarise with `n()`.

```{r}
d %>%
  group_by(ORIGIN) %>%
  summarise(n = n())
```

## Frequencies and Proportions

Can convert frequencies to proportions by dividing through by the total number of rows.

```{r}
d %>%
  group_by(ORIGIN) %>%
  summarise(p = n() / nrow(d))
```

## Central Tendency

Many base functions for describing distribution. Many require `na.rm = TRUE` to remove missing values.

```{r}
no.miss <- c(1, 20, 29.353, .20)
mean(no.miss)
with.miss <- c(1, 20, 29.353, NA)
mean(with.miss)
mean(with.miss, na.rm = TRUE)
```

## Central Tendency

```{r}
d %>%
  filter(!is.na(ARR_DELAY)) %>%
  summarise(
    mean = mean(ARR_DELAY),
    median = median(ARR_DELAY),
    q.5  = quantile(ARR_DELAY, .05),
    q.95 = quantile(ARR_DELAY, .95)
  )
```

## Dispersion

```{r}
# var(d$risk, na.rm = TRUE)  # variance
# sd(d$risk, na.rm = TRUE)  # Standard deviation
# quantile(d$risk, .95)
```

