---
title: "Descriptive Statistics"
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

## Our Data Set

Let's read in the risk data to practice.

```{r}
d <- read.csv("data/risk.csv")
head(d)
```

## Our Data Set

Convert relevant variables to factors.

```{r, results = "hide"}
d$subid <- factor(d$subid)

d$gender <- factor(d$gender, levels = c(0, 1),
                   labels = c("female", "male"))

d$wvcat <- factor(d$wvcat, levels = c(0, 1, 2, 3),
                  labels = c("Unclassifiable", "Individualist", 
                             "Hierarchicalist", "Egalitarian"))
```

## Packages

We'll make use of the `psych` package.

Make sure it's installed on you system and then load it into your workspace.

```{r, eval = F}
install.packages("psych")  # Only if not installed yet
```
```{r, warning = F}
library(psych)  # load it into your workspace
```

# Univariate Analysis

## Frequencies and Proportions

```{r}
my.table <- table(d$gender)
my.table  # Print frequencies

prop.table(my.table)  # Print proportions
```

## Central Tendency

Many base functions for describing distribution. Many require `na.rm = TRUE` to remove missing values.

```{r}
no.miss <- c(1, 20, 29.353, .20)
with.miss <- c(1, 20, 29.353, NA)
mean(no.miss)
mean(with.miss)
mean(with.miss, na.rm = TRUE)
```

## Central Tendency

```{r}
mean(d$risk, na.rm = TRUE)  # The mean
median(d$risk, na.rm = TRUE)  # The median
```

## Dispersion

```{r}
var(d$risk, na.rm = TRUE)  # variance
sd(d$risk, na.rm = TRUE)  # Standard deviation
```

## Conditional Distribution

Sometimes you're interested in knowing the number/proportion of values that meet some condition. This takes two steps. The trick is to work with boolean values, knowing that `TRUE` is treated as `1`, and `FALSE` as `0`.

1. Create boolean vector corresponding to conditional.
2. Treating this as a vector of 1s and 0s, use the likes of `sum()` or `mean()` to compute the sum or proportion of TRUE.

## Conditional Distribution

```{r}
over50 <- d$risk > 50
sum(over50, na.rm = TRUE)
mean(over50, na.rm = TRUE)
prop.table(table(over50))
```

## Some Simple Plots

Try these:

```{r, eval = F}
# Barplot for factors
my.table <- table(d$gender)
barplot(my.table)

# Plots for numeric variables
hist(d$risk)  # Histogram
boxplot(d$risk)  # boxplot
```

## Conditional Distribution

Common condition is the number of missing values, handled using `is.na()`

```{r}
is.miss <- is.na(d$risk)
table(is.miss)
```

## Describe with psych

```{r}
describe(d)
```

# Bivariate Analysis

## Cross-tabs (Factor x Factor)

Add more arguements to `table()`

```{r}
my.table <- table(d$gender, d$wvcat)
my.table  # Print frequencies
prop.table(my.table)  # Cell proportions
```

## Describing by Group (Factor x Numeric)

A simple method is provided by the `psych` package using `describeBy(mydata, group, ...)`

```{r}
describeBy(d[, c("age", "risk")], d$gender)
```

## Multivariate Visualisations (*n* Numeric) 

Again, `psych` provides a great function, `pairs.panels(data)`

```{r}
pairs.panels(d[, c("age", "risk")])
```