---
title: "Descriptive Statistics"
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# Setup

## Packages

We'll make use of the `dplyr` and `psych` packages.

Make sure they're installed on you system and then attach them to your workspace.

```{r, eval = F}
# Install the package from CRAN (if necessary)
install.packages(c("dplyr", "psych"))

# Attach to your workspace
library(dplyr)
library(psych)
```
```{r, include = F}
library(dplyr)
library(psych)
```

## Our Data Set

Let's read in the flight times data to practice.

```{r}
d <- read.csv("data/flight-times.csv")
head(d)
```

# dplyr Summarising and Grouping

## Summarising Data

To summarise information about a data frame, we can use the dplyr function, `summarise()`

```{r}
summarise(d, mean = mean(crs_dep_time))
```

## Splitting data into groups

Often we're interested in splitting our data into groups.

If we plan on using `summarise()`, we can do this with the dplyr function, `group_by()`

For example:

```{r}
by.fn <- group_by(d, origin)
```

Other dplyr functions will now treat `by.fn` as multiple data frames: one for each level of the origin factor.

More to come...

# Univariate Analysis

## Frequencies and Proportions

```{r}
table(d$carrier)  # Print frequencies
table(d$carrier) %>% prop.table()  # Print proportions
```

## Frequencies and Proportions

A dplyr solution: `group_by(factor)` and then `summarise()` with `n()`.

```{r}
d %>%
  group_by(carrier) %>%
  summarise(n = n())
```

## Frequencies and Proportions

and for proportions, divide through by the total number of rows in the data.

```{r}
d %>%
  group_by(carrier) %>%
  summarise(n = n(), p = n / nrow(d))
```

## Test Yourself

Summarise the frequency and proportions of each destination.

Arrange the results from most to least frequent.

Tip: use dplyr functions.

## Central Tendency

Many base functions for describing distribution. Many require `na.rm = TRUE` to remove missing values.

```{r}
no.miss <- c(1, 20, 29.353, .20)
mean(no.miss)
with.miss <- c(1, 20, 29.353, NA)
mean(with.miss)
mean(with.miss, na.rm = TRUE)
```

## Central Tendency

```{r}
mean(d$dep_delay, na.rm = TRUE)  # mean
median(d$dep_delay, na.rm = TRUE)  # Median
```

## Test Yourself

Compute the mean and median of the amount of time spent taxiing ("taxi_out").

## Position

```{r}
min(d$dep_delay, na.rm = TRUE)  # minimum/smallest value
max(d$dep_delay, na.rm = TRUE)  # maximum/largest value
```

## Position

```{r}
quantile(d$dep_delay, na.rm = TRUE)  # Quantiles
quantile(d$dep_delay, .95, na.rm = TRUE)
quantile(d$dep_delay, c(.20, .80), na.rm = TRUE)
```

## Test yourself

Find the values that encapsulate 95% of the departure delay times closest to the median.

## Dispersion

```{r}
range(d$dep_delay, na.rm = TRUE)  # range
var(d$dep_delay, na.rm = TRUE)  # variance
sd(d$dep_delay, na.rm = TRUE)  # Standard deviation
```

## Conditional Distribution

Sometimes you're interested in knowing the number/proportion of values that meet some condition. This takes two steps. The trick is to work with boolean values, knowing that `TRUE` is treated as `1`, and `FALSE` as `0`.

1. Create boolean vector corresponding to conditional.
2. Treating this as a vector of 1s and 0s, use the likes of `sum()` or `mean()` to compute the sum or proportion of TRUE.

## Conditional Distribution

```{r}
sum(d$dep_delay > 60, na.rm = TRUE)
mean(d$dep_delay > 60, na.rm = TRUE)
prop.table(table(d$dep_delay > 60))
```

## Putting it Together with dplyr

Can combine with `summarise()` into a useful table.

```{r}
d %>%
  filter(!is.na(dep_delay)) %>%  # remove rows with missing values
  summarise(
    n = n(),
    n.delayed = sum(dep_delay > 0),
    mean = mean(dep_delay),
    median = median(dep_delay),
    sd = sd(dep_delay),
    q5 = quantile(dep_delay, .05),
    q95 = quantile(dep_delay, .95)
  )
```

## Putting it Together with dplyr

Remember, it's easy to save results as a .csv file.

```{r, eval = F}
x <- d %>%
      filter(!is.na(dep_delay)) %>%  # remove rows with missing values
      summarise(
        n = n(),
        n.delayed = sum(dep_delay > 0),
        mean = mean(dep_delay),
        median = median(dep_delay),
        sd = sd(dep_delay),
        q5 = quantile(dep_delay, .05),
        q95 = quantile(dep_delay, .95)
      )
write.csv(x, "descriptives.csv")
```

## Some Simple Plots

Barplot for factors

```{r}
my.table <- table(d$carrier)
barplot(my.table)
```

## Some Simple Plots

Histogram for numeric variable.

```{r}
hist(d$dep_delay)  # Histogram
```

## Some Simple Plots

Boxplot for numeric variable.

```{r}
boxplot(d$dep_delay)  # boxplot
```

## Some Simple Plots

Boxplot for numeric variable.

```{r}
d %>%
  filter(dep_delay > quantile(dep_delay, .2, na.rm = TRUE) &
         dep_delay < quantile(dep_delay, .8, na.rm = TRUE)) %>%
  select(dep_delay) %>%
  boxplot()
```

# Other Quick Variable Summaries

## Overall Summmaries

Easy base function, `summary(data)`

```{r}
summary(d)
```

## Overall Summmaries

Another is `describe(data)` from the `psych` package.

```{r, warning = F}
describe(d)
```

# Bivariate Analysis

## Cross-tabs (Factor x Factor)

Add more arguements to `table()`

```{r}
table(d$origin, d$carrier)

d %>%
  select(origin, carrier) %>%
  table() %>%
  prop.table()

prop.table(my.table)  # Cell proportions
```

## Cross-tabs (Factor x Factor)

A dplyr solution: `group_by(factor1, factor2, ...)` and then `summarise()` with `n()`.

```{r}
d %>%
  group_by(origin, carrier) %>%
  summarise(n = n())
```

## Describing by Group (Factor x Numeric)

The dplyr approach can be easily extended to multiple groups with `group_by()`

```{r}
d %>% group_by(carrier) %>%
  summarise(n = n(), mean = mean(dep_delay, na.rm = T))
```

## Describing by Group (Factor x Numeric)

Another descriptive method is provided by the `psych` package using `describeBy(mydata, group, ...)`

```{r}
describeBy(d[, c("dep_delay", "taxi_out")], d$carrier)
```

## Multivariate Visualisations (*n* Numeric)

`psych` provides a great function, `pairs.panels(data)`

```{r}
d %>%
  select(crs_dep_time, taxi_out, dep_delay) %>%
  pairs.panels()
```
