---
title: 'Using R: Getting Started'
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# The Plug

## What is R?

A language and environment for statistical computing and graphics.

A serious rival to commercial packages like Matlab, SAS, SPSS, MPlus, etc.

Completely ‘code based’
- We type commands, computer responds (hopefully)
- But, you don’t need to be a computer scientist*

## Why use R?

It’s free and open-source!

For those with no/little experience using software like R (e.g., SPSS users), you’ll quickly learn that R gives you the flexibility to do things with data that you didn’t even think were possible. Put simply, if you can think it, R can do it! The down side, of course, is the learning curve.

For those with experience using similar software/other languages like python, R has its pros and cons compared to these. By-and-large, R is the best at what it promises: statistical analysis and graphics.

# The Warnings

## With Great Power...

Once you being to gain experience with R, you quickly learn that:

- there are a potentially infinite number of ways to achieve the same outcome.
- its power comes from thousands of user-contributed packages, making its mastery endless, and no course or instructor comprehensive.

I will try to focus on fundamental approaches so that you can learn from others. 

## Mental Shift

Using R requires a completely different mental approach to point-and-click software like SPSS.

When you start, it's easy to feel lost.

You do not "see" your data.

If you make a mistake, it can be difficult to locate the problem.

You do not set your analysis up and sift through mountains of output.

I will tend to use code that is explicit, consistent and understandable, to support learning.
 
# Getting started

## Download and Install R

Go to <http://www.r-project.org/>

Click ‘CRAN’, under “Download, Packages”.

Click any link under your country.

Follow links to download latest version on your operating system.

If you are asked to select between 32-bit or 64-bit, you need to match this to your computer. If unsure, take a look at advice online like this <http://www.akaipro.com/kb/article/1616>

Once downloaded, locate the installer. Double click it and follow the prompts to install it on your machine.

## Get RStudio

Great IDE for R.

Go to <http://www.rstudio.com/>

Click RStudio from Products menu.

Click DOWNLOAD RSTUDIO DESKTOP.

Follow links to download latest version on your operating system.

Need to pick 32-bit or 64-bit? Match it to your version of R.

Once downloaded, locate the installer. Double click it and follow the prompts to install it on your machine.

## The RStudio Interface

```{r echo = FALSE, warning = FALSE}
library(png)
library(grid)
img <- readPNG("Rmd/img/RStudio-interface.png")
grid.raster(img)
```

## Create a New Project

1) Open RStudio.
2) Under the File Menu, click **New Project...**
3) Click **New Project**
4) Give your project a name, and decide where to save it on your computer.
5) Click **Create Project**
6) Create a new script by clicking **RScript** under the New File menu (in the File menu), or by clicking the new file logo near the top left (white page with green +).

## Create a New Project

For best practice, you'll want to set up a good folder structure. For example, create the following folders by clicking **New Folder** under the Files tab in one of the panels:

- **R** - use this to store all your R scripts
- **data** - use this to store all your data

## Entering Commands

Commands can be entered directly at the Console prompt `>` by typing and then pressing the enter key.

You can also type them as an R Script in the Source panel, place your mouse cursor on the line and click **Run Line(s)** from the **Code** menu. Shortcut is **ctrl + enter**.

Try run the following commands:

```{r}
1 + 1
2 / 10
```

## Entering Commands

R won’t run until a complete command is entered.

Prompts incomplete commands in the console with `+`.

Try the following:

```{r}
3*
2
```

R sees `3*` and waits for you to multiply 3 by something (in this case, `2`).

## Entering Commands

You can run sections of your code with highlighting in the source panel.

Try to highlight different sections of:

```{r, eval = F}
1 + 2 + 3 + 4
3 - 5
```

Note: R reads commands top to bottom, left to right.

If you want X to happen before Y, X must appear above and/or to the left of Y.

## Annotating your work

Annotate with `#`.

Why?

- Remember what you did and why.
- Allow others to understand your scripts.

Try some examples:

```{r}
# An example of how to annotate in R
3 -
#4  # This number will not be included in the calculation
5  # This number will be included in the calculation
```


## Test yourself

Can you guess the results of the following commands? Run them to check that you're right.

```{r, eval = F}
4 *-
2

4 /
2 +
3 * 1

3*
#2
4

```

Think about some problems

- Order of operations?
- Incomplete commands?

## Saving and opening your scripts

Like anything, save your scripts regularly.

RStudio lets you save and open scripts like files in any other program.

To save: **File/Save…**

- For best practice, save into a scripts folder like **R/**

To open: **File/Open file…**, or use the Files tab in one of your RStudio panels.

## Using packages

Packages are useful collections of functions and data for solving all sorts of problems.

E.g., `psych` for many psychometric analyses, `lme4` package for multilevel modeling, `lavaan` for SEM.

Some current package stats from CRAN...

```{r, echo = F}
types <- c("source", "win.binary", 
           "mac.binary", "mac.binary.mavericks")
 
CRANmirror <- "http://cran.revolutionanalytics.com"
 
pdb <- lapply(types, function(x){
  cran <- contrib.url(repos = CRANmirror, 
                      type = x)
  available.packages(contriburl = cran, type = x)
})
names(pdb) <- types

library(knitr)
kable(sapply(pdb, nrow))

```

## Using packages

Greatly expands the possibilities.

Can be installed just once, but needs to be loaded each session.

Try installing and loading William Revelle's "psych" package:

```{r, eval = F}

# install package (only needed once)
install.packages("psych")

# load package (needed each time)
library(psych)

```

You can see installed and loaded packages in the *Packages* tab of the RStudio panel.

## Getting help

View all supplied documentation with `help.start()`

For getting specific help about a function, run `help(function.name)` or `?function.name`

[Google](www.google.com) is your greatest `R` friend!

Some other great resources:

- [Cookbook for R](http://www.cookbook-r.com/)
- [Quick R](http://www.statmethods.net/)



