---
title: "Using R: Inferential Statistics"
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# t.test()

## Data sets

We'll start with two data sets.

One with a between-subject design

```{r}
b <- read.csv("data/planes.csv")
str(b)
```

## Data sets

One with a within-subject design.

```{r}
w <- read.csv("data/energy_wide.csv")
str(w)
```

## One Sample t-test

One sample *t*-test for a null hypothesis.  
Default settings: null = 0; significance level = .05

```{r}
t.test(b$Distance)
```

## Independent Samples (Between)

`t.test(dv ~ group)`

```{r}
t.test(b$Distance ~ b$Design)
```

## Paired Samples (Within)

`t.test(var1, var2, paired = TRUE)`

```{r}
t.test(w$Running, w$Walking, paired = TRUE)
```

# Statistical Modelling in `R`

## Two-step Process

In general, statistical modelling in R takes two steps.

1. Define and fit your model.
2. Examine the results.

## Defining the Model: The Tilde `~`

In general, statistical models are of the form  
`explain variance in Y with X1, X2, ... Xn`

This is done is `R` using formulas of the form  
`Y ~ X1 + X2 ... Xn`.

Our outcome variable appears to the left of the tilde (`~`) and our predictors appear to the right.

## Defining the Model: Combining Predictors

The predictors can be combined in a number of ways. Here are some common operations:

`Y ~ X1 + X2`  
Predict Y with X1 and X2

`Y ~ X1:X2`  
Predict Y with the interaction of X1 and X2

`Y ~ X1 + X2 + X1:X2` OR `Y ~ X1*X2`  
Predict Y with X1, X2, and their interaction.

# ANOVA - aov()

## One Way (Between)

```{r}
# Read in our data set
d <- read.csv("data/planes.csv")

# 1. Fit the model
fit <- aov(Distance ~ Paper, data = d)

# 2. Explore the results
summary(fit)
```

## Two-or-More Way (Between)

```{r}
# Read in our data set
d <- read.csv("data/fishing.csv")

# 1. Fit the model
fit <- aov(distance ~ rod*line*sinker, data = d)

# 2. Explore the results
summary(fit)
```

## Two-or-More Way (Between)

```{r}
# Two-variable plot
plot(d$sinker, d$distance)
```

## Two-or-More Way (Between)

```{r}
# Three-variable plot
interaction.plot(d$line, d$sinker, d$distance)
```

## Within Subjects

```{r}
d <- read.csv("data/energy.csv")
fit <- aov(energy ~ activity + Error(Subject / activity), data = d)
summary(fit)
```

## Mixed

# Correlation

# Regression - lm()

## Simple Linear Regression

## Multiple Regression

## Hierarchical Regression

## Logistic Regression

## Standardized Coefficients




