---
title: "Using R: Plotting with ggplot2"
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = F, warning = F}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../../')
```

# Setup

## Packages

We'll be using the ggplot2, tidyr, and dplyr packages.

```{r, eval = F}
# Install the package from CRAN (if necessary)
install.packages(c("dplyr", "tidyr", "ggplot2"))

# Attach to your workspace
library(dplyr)
library(tidyr)
library(ggplot2)
```
```{r, include = F}
library(dplyr)
library(tidyr)
library(ggplot2)
```

# ggplot2

## What is ggplot2 and why use it?

In general, a good reason to use R is its strong graphics capabilities.

`ggplot2` is "a plotting system for R, based on the grammar of graphics".

It enables you to produce professional looking, publishable quality, and highly modifiable, graphs.

## ggplot2 Basics

In general, ggplot2 builds a plot using three components:

- A data frame
- geoms, which are the visual elements that represent your data
- A coordinate system

## ggplot2 Warning

`ggplot2` is incredibly powerful, and remember what comes with great power!

You will continue searching for solutions for a long time, perhaps forever.

But the key is to understand the structure, and the rest will follow.

## ggplot2 Cheat Sheet

There's a great cheat sheet online.

Don't hesitate to print it out and stick it on your wall!

# Graphs with ggplot2

## Basic Structure

Let's read in a data set:

```{r}
d <- read.csv("data/flight-times.csv")
head(d)
```

## Basic Structure

We start to build a graph with `ggplot(data.frame, aes(variables, ...))`

```{r, eval = F}
ggplot(d, aes(carrier))
```

This just sets up our data. Next we need to add geoms.

## Basic Structure

We now have the data prepared.

Next, using `+`, we need to add layers and elements.

For example, we should add geom layer (function) that will represent our data as a visual element.

Note that geoms create a default coordinate system.

## Basic Structure

```{r}
ggplot(d, aes(carrier)) + geom_bar()
```

## The Basics

```{r, warning = F}
ggplot(d, aes(dep_delay)) +
  geom_density()
```

## The Basics

Extending to more variables is easy:

```{r, warning = F}
ggplot(d, aes(x = dep_time, y = dep_delay)) + geom_point()
```

## The Basics

You can also save the graph as a variable and grow it.

```{r, warning = F}
p <- ggplot(d, aes(x = dep_time, y = dep_delay))
p <- p + geom_point()
p
```

# Geoms - One Variable

## Bar (for Discrete variables)

```{r, warning = F, message = F}
ggplot(d, aes(carrier)) + geom_bar()
```

## Density

```{r, warning = F, message = F}
ggplot(d, aes(dep_delay)) + geom_density()
```

## Histogram

```{r, warning = F, message = F}
ggplot(d, aes(dep_delay)) + geom_histogram()
```

# Geoms - Two Variables

## Jitter (Discrete X, Discrete Y)

```{r, warning = F, message = F}
ggplot(d, aes(carrier, origin)) + geom_jitter()
```

## Bar (Discrete X, Continuous Y)

```{r, warning = F, message = F}
d %>% group_by(carrier) %>%
  summarise(mean.delay = mean(dep_delay, na.rm = T)) %>%
  ggplot(aes(carrier, mean.delay)) + geom_bar(stat = "identity")
```

## Boxplot (Discrete X, Continuous Y)

```{r, warning = F, message = F}
ggplot(d, aes(carrier, dep_delay)) + geom_boxplot()
```

## Boxplot (Discrete X, Continuous Y)

```{r, warning = F, message = F}
d %>% filter(dep_delay < quantile(dep_delay, .95, na.rm = TRUE)) %>%
  ggplot(aes(carrier, dep_delay)) + geom_boxplot()
```

## Violin (Discrete X, Continuous Y)

```{r, warning = F, message = F}
d %>% filter(dep_delay < quantile(dep_delay, .95, na.rm = TRUE)) %>%
  ggplot(aes(carrier, dep_delay)) + geom_violin()
```

## Points (Continuous X, Continuous Y)

```{r, warning = F, message = F}
ggplot(d, aes(dep_time, dep_delay)) + geom_point()
```

## Smoothed Conditional Mean (Continuous X, Continuous Y)

```{r, warning = F, message = F}
ggplot(d, aes(dep_time, dep_delay)) + geom_smooth()
```

## Smoothed Conditional Mean (Continuous X, Continuous Y)

```{r, warning = F, message = F}
ggplot(d, aes(dep_time, dep_delay)) + geom_smooth(method = "lm")
```

## Points and Regression Line (Continuous X, Continuous Y)

```{r, warning = F, message = F}
ggplot(d, aes(dep_time, dep_delay)) + geom_point() +
  geom_smooth(method = "lm")
```

## 2D Bins (Continuous Bivariate)

```{r, warning = F, message = F}
ggplot(d, aes(day_of_month, taxi_out)) + geom_bin2d()
```

## 2D Density (Continuous Bivariate)

```{r, warning = F, message = F}
ggplot(d, aes(day_of_month, taxi_out)) + geom_density2d()
```

## Line (Continuous Function)

```{r, warning = F, message = F}
d %>% group_by(day_of_month) %>% summarise(n = n()) %>%
ggplot(aes(day_of_month, n)) + geom_line()
```

## Error Bars

```{r, warning = F, message = F}
x <- d %>% group_by(carrier) %>%
        summarise(n    = n(),
                  mean = mean(dep_delay, na.rm = TRUE),
                  se   = sd(dep_delay, na.rm = TRUE)/sqrt(n))
```

## Error Bars

```{r, warning = F, message = F}
ggplot(x, aes(carrier, mean, ymin = mean - se, ymax = mean + se)) +
    geom_errorbar()
```

## Bars with Error Bars

```{r, warning = F, message = F}
ggplot(x, aes(carrier, mean, ymin = mean - se, ymax = mean + se)) +
    geom_bar(stat = "identity") + geom_errorbar()
```

## Bars with Error Bars

```{r, warning = F, message = F}
ggplot(x, aes(carrier, mean, ymin = mean - se, ymax = mean + se)) +
    geom_bar(stat = "identity") + geom_errorbar(width = .25)
```

## Text

```{r, warning = F, message = F}
ggplot(x, aes(carrier, mean, label = carrier)) +
  geom_text()
```

## Text

```{r, warning = F, message = F}
ggplot(x, aes(carrier, mean, label = carrier)) +
  geom_bar(stat = "identity") +
  geom_text()
```

## Text

```{r, warning = F, message = F}
ggplot(x, aes(carrier, mean, label = carrier)) +
  geom_bar(stat = "identity") +
  geom_text(aes(vjust = ifelse(sign(mean) > 0, -1, 2)))
```

## Text

```{r, warning = F, message = F}
x <- x %>% mutate(ci.upper = mean + se * 1.96,
                  ci.lower = mean - se * 1.96,
                  zero = ci.lower < 0 & ci.upper > 0,
                  flag = ifelse(zero, "", "*"))
```

## Text

```{r, warning = F, message = F}
ggplot(x, aes(carrier, mean, label = flag)) +
  geom_bar(stat = "identity") +
  geom_text(aes(vjust = ifelse(sign(mean) > 0, -1, 2)))
```

# Three Variables

## Scales

Scales apply an aesthetic to geoms.

The main scales you'll work with are:

- shape
- size
- fill
- color

## Shape (A Third Discrete)

```{r, warning = F, message = F}
d %>% filter(carrier %in% c("AA", "B6", "DL") & dep_delay < 15) %>%
  ggplot(aes(dep_delay, arr_delay, shape = carrier)) + geom_point()
```

## Size (A Third Continuous)

```{r, warning = F, message = F}
read.csv("data/physical.csv") %>%
  ggplot(aes(Thigh, Height, size = Mass)) + geom_point()
```

## Fill  (A Third Discrete)

```{r, warning = F, message = F}
d <- read.csv("data/planes.csv")
d %>% group_by(Paper, Angle) %>%
  summarise(Distance = mean(Distance)) %>%
  ggplot(aes(Paper, Distance, fill = Angle)) +
  geom_bar(stat = "identity")
```

## Fill (A Third Discrete)

```{r, warning = F, message = F}
d %>% group_by(Paper, Angle) %>%
  summarise(Distance = mean(Distance)) %>%
  ggplot(aes(Paper, Distance, fill = Angle)) +
  geom_bar(stat = "identity", position = "dodge")
```

## Colour

```{r, warning = F, message = F}
read.csv("data/physical.csv") %>%
  ggplot(aes(Thigh, Height, size = Mass, colour = Mass)) +
  geom_point()
```

# Additional Features

## Labels

The following change the major labels on your graphs:

- `+ ggtitle("Graph Title")`
- `+ xlab("X axis label")`
- `+ ylab("Y axis label")`
- `+ labs(title = "", x = "", y = "")`

## Faceting

Faceting allows you to create panels of subplots plots based on a discrete variable.

```{r, warning = F, message = F}
read.csv("data/planes.csv") %>%
  group_by(Paper, Angle) %>%
  summarise(Distance = mean(Distance)) %>%
  ggplot(aes(Paper, Distance)) + geom_bar(stat= "identity") +
  facet_grid(. ~ Angle)
```

## Faceting

```{r, warning = F, message = F}
read.csv("data/planes.csv") %>%
  group_by(Paper, Angle) %>%
  summarise(Distance = mean(Distance)) %>%
  ggplot(aes(Paper, Distance)) + geom_bar(stat= "identity") +
  facet_grid(Angle ~ .)
```

## Faceting

```{r, warning = F, message = F}
read.csv("data/planes.csv") %>%
  group_by(Paper, Angle, Design) %>%
  summarise(Distance = mean(Distance)) %>%
  ggplot(aes(Paper, Distance)) + geom_bar(stat = "identity") +
  facet_grid(Angle ~ Design)
```

## Themes

Themes adjust overal aesthetic elements.

```{r}
read.csv("data/flight-times.csv") %>% 
  ggplot(aes(carrier)) + geom_bar() +
  theme_classic()
```

## Saving

`ggsave()` saves the last plot you created.

```{r, eval = F}
ggsave("plot.png", width = 5, height = 5)
```

# Putting it Together

## Challenge

```{r}
d <- read.csv("data/flight-times.csv")
```
Using data for carriers "AA" and "DL", flights leaving after 5AM and departing between 0 and 20 minutes late, create a scatter plot of the departure delay (`dep_delay`) against the scheduled departure time (`crs_dep_time`). Add a loess line, and colour the points by `carrier`. Make sure appropriate axis labels are used, and a classic theme. Save the graph as a 5 x 3 jpeg file.

Bonus: use a single pipeline to create the graph.

## Challenge

```{r, eval = F}
d %>% filter(carrier %in% c("AA", "DL") &
               crs_dep_time > 500 &
               dep_delay > 0 & dep_delay < 20) %>%
  ggplot(aes(crs_dep_time, dep_delay, colour = carrier)) +
    geom_point() +
    geom_smooth() +
    labs(x = "Scheduled Departure Time",
         y = "Departure Delay") +
    theme_classic()
ggsave("plot.jpeg", width = 5, height = 3)
```
