---
title: 'Using R: programming challenge'
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3
)
```

## Motivation

Data collected........... check!

Data cleaned............. 

Big job when cleaning is to check whether there are any values outside of a valid range.

## Example

We have multiple columns, each with different possible values.

We can't look at every cell.

What to do?

## Challenge

Say the following are valid values for variables in the `iris` data:

- `Sepal.Length`  4.5 to 8
- `Sepal.Width`   1 to 5
- `Petal.Length`  1.1 to 6.8
- `Petal.Width`   0 to 2
- `Species`       "setosa" and "versicolor"

Use programming methods to check how many non-valid values are in each column.

You will need conditional execution, functions, lists, and iteration.

## Bonus

Create plots that helps you to vizualise the results in a useful way.

Examples:

- Bar plot of the number of invalid values per variable.
- Tile plot showing row numbers of invalid values for each value.

## One answer

```{r}
d <- iris

# Function to return number of values
# in a vector that are not valid
#
# Args:
#   x Vector of values to check
#   valid Vector of valid values.
#         2-elements (low and high) if numeric,
#         more if factor.
not_valid <- function(x, valid) {
  if (is.numeric(x)) {
    valid <- x >= valid[1] & x <= valid[2]
  } else {
    valid <- x %in% valid
  }
  
  sum(!valid)
}
```

## One answer

```{r}
valid_list <- list(
  Sepal.Length = c(4.5, 8),
  Species     = c("setosa", "versicolor")
)


not_valid(iris$Sepal.Length, c(4.5, 8))
not_valid(iris$Species, c("setosa", "versicolor"))

vars <- names(valid_list)
names(vars) <- vars

purrr::map_int(vars, ~ not_valid(d[[.]], valid_list[[.]]))
```

