---
title: 'Using R: conditional execution'
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3
)
```

## `if`

We often want to run code only `if` a certain condition is `TRUE`.

For example, we want to be warned if there are invalid scores in a variable.

This is achieved with an `if` statement:

```{r, eval = F}
if (condition) {
  # code to run if the condition results in TRUE
}
```

Note, `condition` is anything that evalutes to `TRUE` or `FALSE`.

## `if`

Let's see how it works:

```{r}
if (TRUE) {
  print("Hello")
}

if (FALSE) {
  print("Hello")
}
```

## `if`

```{r}
x <- 1
y <- 2

if (x == 1) {
  print("Hello")
}

if (x == 1 & y != 2) {
  print("Hello")
}
```

## Test Yourself

Create two vectors: one numeric and the other a character vector.

Write two if statements to check whether each vector `is.numeric()`.

If it is, then compute the `mean()`.

# Logical Operators and functions

## Logical operators and functions

A reminder of the logical operators and functions you're likley to use a lot:

- `==`, `!=`, `<`, `>`, `<=`, `>=`
- `&`, `|`
- `!`
- `all()`, `any()`
- `%in%`

## Test yourself

Create the following vectors:

```{r}
x <- 1:10
y <- 21:30
```

Print an error message if x contains a particular value (e.g., 8) and all the values in y are a particular sign (e.g., positive).

# Else

## Else?

Say you want to do one thing if your condition is `TRUE`, but another thing if it is `FALSE`.

How would you achieve this?

## Else?

We could write exclusive if statements.

Even better, we could save our condition as a boolean variable and write two if statements that use variable and !variable.

Or...

## `if else`

If our condition is `FALSE`, we can run different code with `else` as follows:

```{r, eval = FALSE}
if (condition) {
  # code to run if condition is TRUE
} else {
  # code to run  if condition is FALSE
}
```

## `if else`

An example:

```{r}
x <- "a"

if (x == "b") {
  print("x is b")
} else {
  print("x is not b")
}
```

## Test yourself

Create a variable `x`. E.g., `x <- 2`.

Write an if statement that checks whether `x` is numeric and, if so, prints "number". Else, print, "other".

Try for different values of `x`.

## Test yourself

Create two vectors, one numeric and one character.

Write if-else statements to check whether each vector `is.numeric()`.

If so, compute the vector mean; else show the frequencies using `table()`.

## `if else if else`

We can add more conditionals as follows:

```{r, eval = FALSE}
if (condition 1) {
  # code to run if condition 1 is TRUE
} else if (condition 2) {
  # code to run if condition 2 is TRUE
} else if (condition 3) {
  # code to run if condition 3 is TRUE
}
...
} else {
  # code to run if no conditions are TRUE
}
```

## Test Yourself

Create an object `x` with some value (e.g., `NA`, `10`, `"a"`)

Create a condition to:

- `print("Missing")` if `x` is a missing value.
- Else, `print("Number")` if it is a number.
- Otherwise, `print("Other")`

Trial your conditional execution with different values for `x`.

# Challenge

## Challenge

Save and work with a copy of the `mtcars` data.

For example sake, say that the range of valid values for the variable `qsec` is 10 to 20.

Write an `if` statement that checks for any invalid values in this variable.

If there are, remove the rows associated with those values and print out a message that says how many rows were removed.

If there are not, print out a message that says all values are invalid.