---
title: 'Using R: conditional execution'
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3
)
```

## `if`

We often want to run code only `if` a certain condition is `TRUE`.

For example, we want to be warned if there are invalid scores in a variable.

This is achieved with an `if` statement:

```{r, eval = F}
if (condition) {
  # code to run if the condition results in TRUE
}
```

Note, `condition` is anything that evalutes to `TRUE` or `FALSE`.

## `if`

Let's see how it works:

```{r}
if (TRUE) {
  print("Hello")
}

if (FALSE) {
  print("Hello")
}
```

## `if`

```{r}
x <- 1
y <- 2

if (x == 1) {
  print("Hello")
}

if (x == 1 && y != 2) {
  print("Hello")
}
```

## `if else`

If our condition is `FALSE`, we can run different code with `else` as follows:

```{r, eval = FALSE}
if (condition) {
  # code to run if condition is TRUE
} else {
  # code to run  if condition is FALSE
}
```

## `if else`

An example:

```{r}
x <- "a"

if (x == "b") {
  print("x is b")
} else {
  print("x is not b")
}
```

## `if else if else`

We can add more conitionals as follows:

```{r, eval = FALSE}
if (condition 1) {
  # code to run if condition 1 is TRUE
} else if (condition 2) {
  # code to run if condition 2 is TRUE
} else if (condition 3) {
  # code to run if condition 3 is TRUE
}
...
} else {
  # code to run if no conditions are TRUE
}
```

## Test Yourself

Create an object `x` with some value (e.g., `NA`, `10`, `"a"`)

Create a condition to:

- `print("Missing")` if `x` is a missing value.
- Else, `print("Number")` if it is a number.
- Otherwise, `print("Other")`

Trial your conditional execution with different values for `x`.