---
title: 'Using R: writing functions'
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3
)
```

## Functions for everyone

Functions don't just come in R or packages that we install.

We can write our own functions.

But what's the point?

## Why write our own functions

Most of the time, we write functions to replace the need to copy and paste.

For example, instead of constantly copying and pasting the code needed to compute the mean of different vectors (e.g., computing sum, compute length, divide through), we can use `mean()`.

## Why write our own functions

The major advantages of using functions instead of copy and paste is that:

- We avoid forgetting to make changes throughout large code chunks.
- If we need to change our code, we just need to do it in one place.
- Keeps our code clean and legibile.

## How to create and use a function

We create (**declare**) a function by:

`func_name <- function(arg1, arg2, ...) { code }`

1. Declare a function name.
2. Assign it (`<-`) to `function() {}`.
3. Place arguments whose values can be changed inside the parentheses `()`.
4. Place code to be executed inside the curly braces `{}`.
5. Ensure we return a value in the last code line or with `return()`.

## How to create and use a function

We then use (**call**) it by:

`func_name(val1, val2, ...)`

1. Call it's name followed by parentheses `()`.
2. Providing values for the arguments inside the parentheses `()`, separated by commas.

## How to create and use a function

Example:

```{r}
# Function to add one 
add_1 <- function(x) {
  x + 1
}

add_1(4)

a <- c(1, 2, 3)
add_1(a)
```

## How to create and use a function

Example:

```{r}
# Count the number of values equal to something
num_of <- function(vals, equal_to) {
  sum(vals == equal_to)
}

x <- c(1, 1, 3)
y <- c("a", "a", "b")
num_of(x, 1)
num_of(x, 88)
num_of(y, "b")
num_of(y, "z")
```

## Test yourself

Write a function that divides a number by 2.

Next, adjust the function so that you can divide by any number instead of just 2.

# Arguments

## Arguments

Arguments give functions flexibility.

We substitute arguments for values when we call our function.

So far, we've done this based on the **order** of the arguments.

## Arguments by name

It's possible, however, to reference arguments by name:

```{r}
x <- c(1, 1, 3)
num_of(x, 1)
num_of(equal_to = 1, vals = x)
```

## Argument defaults

We can also set argument defaults when we declare the function.

When a default exists, we don't need to include a value for that argument when we call the function.

```{r}
divide_by <- function(val, divisor = 2) {
  val / divisor
}

divide_by(4)
divide_by(4, 3)
divide_by(4, divisor = 5)
```

## Test yourself

Write a function that does some arithmetic with at least three arguments.

Provide default values for the second and third arguments.

Run your function by providing values for the first and third arguments.

# Return

## Return

Functions must also return something. By default, this will be the result of the last line of code. However, we can do this explicitly with `return()`.

```{r}
fun_1 <- function(x) {
  x
  x * 2
}

fun_2 <- function(x) {
  return (x)
  x * 2
}

fun_1(10)
fun_2(10)
```

## Return

Using `return()` is particularly useful when combined with conditional statements. For example, try the following:

```{r}
add <- function(x, y) {
  if (is.na(x)) {
    return (y)
  } else if (is.na(y)) {
    return (x)
  } else {
    return(x + y)
  }
}
```

## Free yourself with source()

As we start to write many functions, it becomes important to keep them organised and in a way that makes it possible to transport/share them across projects.

A good way to do this is to place your functions in separate function scripts and then use `source()` to read the script in.

For example...

1. Place your functions in a new script (e.g., called `"functions.R"`).
2. Save this script in the "R/" folder.
3. Open a new script, and place `source("R/functions.R")` at the top.
4. Run this to be able to use your functions.

## Test yourself

Write a function to count how many numeric values of a vector are outside an acceptable range.

For example, to check whether any values in a vector of proportions are outside the range of zero to one.

Place this function in a function R script and use `source()` to declare and use it in another script.