---
title: 'Using R: iteration'
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3
)
```

## What is Iteration?

Iteration is a critical component of programming that involves executing certain code multiple times.

For example, we use iteration to run a check on every column or row of our data, or to run a list of regression models.

Like writing functions, iteration helps us to cut down on the need to copy and paste code.

## Example without iteration

How would you calculate the standard deviation of every column in the `mtcars` data set?

## Example without iteration

You'd probably start with something like:

```{r}
sd(mtcars$mpg)
sd(mtcars$cyl)
sd(mtcars$disp)
sd(mtcars$hp)
```

etc.

# for loops

## `for` loops

`for` loops are a typical way to do iteration in many programming langauages.

We write a `for` loop in R as follows:

```{r, eval = F}
for (i in sequence) {
  # code to run here making use of `i`
}
```

## Quick examples

```{r, eval = F}
x <- 1:5
for (i in x) {
  print("Hi")
}
```

## Quick examples

```{r, eval = F}
x <- 1:5
for (i in x) {
  print(i)
}
```

## Quick examples

```{r, eval = F}
x <- 1:5
for (i in x) {
  print(i + 1)
}
```

## The `for` loop

At a minimum, a `for` loop has two main components:

- the `sequence`
- the code to execute for each value in the `sequence`

The sequence is usually a vector or list, and we execute the code once for each value in the sequence.

That is, we run the loop for the first sequence element, then the second, all the way up to the *n*^th^.

## The `for` loop

We access the *i*^th^ element of the sequence in a given iteration by way of:

```{r, eval = F}
i in sequence
```

This means we can use `i` within the `for` code to refer to the *i*^th^ element of the sequence.

Aside, note that `i` can be any valid variable name.

## The `for` loop

Here's a basic example:

```{r}
for(i in 1:5) {
  print(i)
}
```

## The `for` loop

Another

```{r}
for(i in 1:5) {
  x <- i / 2
  print(x)
}
```

## Test yourself

Create a vector of names.

Write a `for` loop that prints each name, one at a time.

## Test yourself

Try writing a `for` loop with a list as a sequence rather than a vector.

## Sequence variations

Recall that a data frame is a list (though not very effective):

```{r}
for (col in mtcars) {
  print(sd(col))
}
```

## Sequence variations

How about this:

```{r}
for(var in names(mtcars)) {
  dat <- mtcars[[var]]
  cat(var, ":", sd(dat), "\n")
}
```

## Saving the results

`for` loops don't need to return anything, but generally we want them to.

When saving results, it's best practice to declare an (empty) object first, then save to it.

## Saving the results

For example:

```{r}
sds <- vector("double", ncol(mtcars))
names(sds) <- names(mtcars)
sds

for(var in names(mtcars)) {
  sds[[var]] <- sd(mtcars[[var]])
}
sds
```

# Challenge

## Challenge

Use a loop to compute the median of every column in the `airquality` data set.

Save the results in a named vector.

Plot the results in a `barplot()`.

Bonus: `order()` the results from lowest to highest and then plot.


## Challenge

Answer:

```{r, fig.width=7}
meds <- vector("double", ncol(airquality))
names(meds) <- names(airquality)
for(var in names(airquality)) {
  meds[[var]] <- median(airquality[[var]], na.rm = TRUE)
}
meds <- meds[order(meds)]
barplot(meds, ylab = "Median")
```

# Beyond for loops

## Going Beyond `for`

Understanding `for` loops is a foundational programming skill.

In R, however, you have a number of unique and superior choices.

Most prominently, there are:

- \*apply functions in base R
- map\* functions in the purrr package.

## Going Beyond `for`

Using \*apply or map\* functions involves writing the relevant "`iteration_function()`" in the following format (literally or conceptually):

```{r, eval = F}
iteration_function(sequence, function)
```

- The function is applied/mapped to each sequence element. No need to write `for i in sequence`.
- They always return the iterated function results. No need to create empty objects.

# *apply()

## `lapply()`

`lapply()` is one of the most frequently used of the \*apply() functions. It returns the results of the function as a **list**:

```{r}
lapply(mtcars, sd)
```

## `lapply()`

```{r, eval = F}
lapply(mtcars, sd)
```

Things to note:

- No parentheses: `sd` not `sd()`
- Value for each iteration is used as the **first** function argument
- Sequence names transferred to result
- We can immediately save it to a variable:

```{r}
x <- lapply(mtcars, sd)
x$hp
```

## `lapply()`

Additional function arguments can be passed after the function:

```{r}
lapply(airquality, sd, na.rm = TRUE)
```

## Test Yourself

Use `lapply()` to:

- calculate the range of every column in the `mtcars` data set.
- save the median of every column in the `airquality` data set, and print only the `Ozone` results

## Test Yourself

Answers

```{r, eval = F}
# 1
lapply(mtcars, range)

# 2
x <- lapply(airquality, median, na.rm = TRUE)
x$Ozone
```

## `sapply()`

`sapply()` is exactly the same as `lapply()` except that it attempts to **simplify** the results into a vector or matrix:

```{r}
sapply(mtcars, range)
```

[Beware of performance issues](http://stackoverflow.com/questions/38649411/r-speed-up-the-for-loop-using-apply-or-lapply-or-etc/38649860#38649860).

## Test Yourself

Use `sapply()` to compute the quantiles (`quantile()`) of all `mtcars` columns and return the results as a matrix.

Bonus: add a pipe (`%>%`) into your code.

## Test Yourself

Answer:

```{r, message=F}
library(dplyr)
mtcars %>% sapply(quantile)
```

# purrr::map*()

## purrr package

The purrr package "makes your pure functions purr".

Install and attach:

```{r, message=F}
# install.packages("purrr")
library(purrr)
```

## `map()`

The primary purrr function is `map()`.

It takes a sequence and function, and returns a list, just like `lapply()`:

```{r}
mtcars %>% map(mean)
```

## Test Yourself

Use `map()` to find the median of each `mtcars` column.

Save the results and then subset it to get the median for `hp` and `mpg`.

Write the `lapply()` equivalent.

## `map_*()`

For simplified output, there are a variety of map functions:

- `map_lgl()`/`map_int()`/`map_dbl()`/`map_chr()`/`map_df()`

```{r}
airquality %>% map_dbl(mean, na.rm = TRUE)

airquality %>% map_df(range, na.rm = TRUE)
```

## Test Yourself

Using a varient of `map()` to simplify the output, check which columns of the `iris` data set are numeric with the function `is.numeric`.

Bonus: use the results to calculate the total number of numeric columns.

# Custom functions

## Adding custom functions

There are two ways we can add custom functions into the \*apply and map\*:

1. Create the function then use its name.
2. Use an anonymous function

## Using a created function

When we write our own functions, we can use them like any other in the iteration functions covered:

```{r}
# Function to normalise a vector to range between 0 and 1
normalise <- function(vector) {
  (vector - min(vector)) / (max(vector) - min(vector))
}

mtcars %>% map_df(normalise) %>% head()
```

## Anonymous functions

Anonymus functions are functions that do not exist in the environment, and are written directly into another function like `lapply()`. 

For `*apply()` functions, we do the following:

```{r, eval = FALSE}
*apply(sequence, function(i) {
  # Code using `i`
})
```

For `map*()` functions, we do the following:

```{r, eval = FALSE}
map(sequence, ~ # code using `.`)
```

## Examples

Here's an anonymous version of the normalise function:

```{r}
mtcars %>% lapply(function(i) {
  (i - min(i)) / (max(i) - min(i))
})
```


## Examples

Here's an anonymous version of the normalise function:

```{r}
mtcars %>% map_df(~ (. - min(.)) / (max(.) - min(.)))
```

# Challenge

## Challenge

`split()` the `mtcars` data based on the number of car cylinders into a list as follows:

```{r, eval = F}
mtcars %>%
  split(.$cyl)
```

Continue the pipe and regress `hp` on `wt` and `mpg` for each split. Try using `lapply()` and `map()`.

Bonus: continue the pipe again and obtain the `summary()` for each fitted model.

Bonus Bonus: A lm summary is a list containing `"r.squared"`. Can you extract it for each model?

## Challenge

One answer

```{r}
mtcars %>% 
  split(.$cyl) %>% 
  map(~ lm(hp ~ wt + mpg, data = .)) %>% 
  map(summary) %>% 
  map_dbl("r.squared")  # OR lapply(function(i) i$r.squared)
```

## Challenge

A bonus for you to try

```{r, eval = FALSE}
mtcars %>%
  split(.$cyl) %>%
  map(~ lm(hp ~ wt + mpg, data = .)) %>%
  map(summary) %>%
  map("coefficients")
```
