---
title: 'Using R: lists'
author: "Simon A Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
# require(knitr, quietly = TRUE)
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  #root.dir = '../../../../',
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3
)
```

## list() v c()

A `list()` is **like** a vector (`c()`) because:

- It is a sequence with a particular length.
- It can be named.

A `list()` is **different** from a vector (`c()`) because:

- It can be heterogeneous.
- Contain anything, even other lists!

## Creating lists

We create a list with `list()`, which we use similarly to `c()`:

```{r}
x <- list(1, 2, 3)
class(x)
x
```

## Creating lists

Each element can be anything.

```{r}
x <- list(1, "a", TRUE, c(1, 2), list("b", 2))
str(x)
```

## Naming lists

List elements can be named

```{r}
x <- list(a = 10, b = TRUE, "Hello")
names(x)[3] <- "c"
x
```

## Indexing lists

Indexing can be done in a few ways, which each have important differences:

- `[`
- `[[`
- `$` 

## Indexing lists

`[` extracts a subset of the list, and returns a list.

```{r}
x[c(1, 3)]
x[c("a", "b")]
class(x[c(1,3)])
```

## Indexing lists

`[[` extracts a single element, and returns the list element, whatever it is.

```{r}
x[[1]]
x[["c"]]
class(x[[2]])
```

## Indexing lists

`$` is a convenience method for `[[` when the list is named.

```{r}
x$a
x$b
class(x$c)
```

## Changing a list

Use any of the indexing methods to add a new list element or change an existing one.

```{r}
x$a  <- 27
x[4] <- c("I'm new")
x$d  <- list(TRUE, FALSE)
```

## Changing a list

```{r}
x
```

## Test Yourself

Create a named list with the following objects:

- A numeric scalar
- A numeric vector

Add an additional object which is an unnamed list containing:

  - A numeric scalar
  - A numeric vector
  
Add the numeric scalar in the main list to the numeric scalar in the nested list.

## Regression example

Let's test a multiple regression model with and without an interaction term, and store all the results in a list, `lms`.

```{r}
# Import data; store the models
d <- read.csv("data/physical.csv")
lms <- list()
lms$mods <- list(
  main = "Mass ~ Fore + Bicep",  # main effects only
  int  = "Mass ~ Fore * Bicep"   # interaction included
)
str(lms)
```

## Regression example

Now, let's fit the models.

```{r}
lms$fits <- list(
  main = lm(lms$mods$main, data = d),
  int  = lm(lms$mods$int,  data = d)
)
```

Examine the list so far.

## Regression example

Summarise results for whichever fit we're interested in:

```{r}
summary(lms$fits$int)
```

## Test yourself

Import the `planes.csv`.

Create a list and fill it with t-tests for differences in `Distance` as a function of each of the three categorial variables (`Paper`, `Angle`, and `Design`).

Be able to use your list to access the results for whichever categorical variable you might be interested in.

## A note about data frames

Data frames are lists with certain restrictions/features:

- Each column is an object in the list.
- Each column must be homogeneous.
- Each column must be the same length.
