---
title: 'Using R: ggplot2 facets'
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 3,
  fig.width = 3,
  fig.align = "center"
)
```

## Facets

Facets refer to subpanels that are created by dividing the plotting area. For example:

```{r, echo = F, fig.width=8}
library(ggplot2)
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  facet_wrap(~ cyl) +
  ggtitle("Breakdown by number of cylinders")
```

## When/how to use facets

We typically want to use facets with one or two categorical variables.

There are two facet functions available to us:

```{r, eval = F}
facet_grid()

facet_wrap()
```

Like all other ggplot2 functions, we use these by adding them to a sequence of plot functions. E.g.,

```{r, eval = F}
ggplot() + geom_*() + facet_wrap()
```

# facet_wrap()

## facet_wrap()

`facet_wrap()` is used to layout 1D panels splits in a 2D manner. For this we use `~ var_to_wrap`:

```{r, fig.width=5, warning=F}
ggplot(airquality, aes(x = Ozone, y = Temp)) + geom_point() +
  facet_wrap(~ Month)
```

## Test yourself

Use `facet_wrap()` to create a separate scatter plot for each `Day` instead of month.

# facet_grid()

## facet_grid()

`facet_grid()` lays out panels in a grid. We define which variable to split on rows, columns, or both.

The first argument is used to define the panels in the form:

```{r, eval = F}
facet_grid(var_to_split_rows ~ var_to_split_cols)
```

## facet_grid()

Example:

```{r, fig.width=8}
ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point() +
  facet_grid(am ~ cyl)
```

## facet_grid()

To NOT divide rows or columns, use `.`:

```{r}
ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point() +
  facet_grid(am ~ .)
```

## Test yourself

Split the plot into columns only.

# Multiple facets on same axis

## Multiple variables

In both facet functions, we can add (`+`) multiple variables to facet by. For example:

```{r}
ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point() +
  facet_wrap(~ cyl + am)
```

## Test yourself

Create a scatter plot of two continuous variables from the `mtcars` dataset.

Use `facet_wrap()` and `facet_grid()` to divide the plot depending on the number of gears (`gear`) and/or the number of carbuetors (`carb`) a car has.

Try various combinations of facets the check that you can obtain what you think you can.

# Challenge

## Challenge 1

Create a facetted `geom_density` of every variable in the `mtcars` data set:

```{r, echo = F, message = F, fig.height=4, fig.width=7}
library(dplyr)
library(tidyr)

mtcars %>%
  gather() %>% 
  ggplot(aes(x = value)) +
    geom_density() +
    facet_wrap(~ key, scales = "free")
```

## Challenge 2

Using the `iris` data, create facetted `geom_histogram()` of all continuous variables (all but `Species`).

```{r, echo = F, message = F, fig.height=4, fig.width=7}
library(dplyr)
library(tidyr)

iris %>%
  select(-Species) %>% 
  gather(var, val) %>% 
  separate(var, into = c("part", "measure")) %>% 
  ggplot(aes(x = val)) +
    geom_histogram() +
    facet_grid(part ~ measure)
```
