---
title: 'Using R: ggplot2 geoms and aesthetics'
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 3,
  fig.width = 3,
  fig.align = "center"
)
```

## Getting Started

(Install and) Load the `ggplot2` package:

```{r}
# install.packages("ggplot2")
library(ggplot2)
```

A ggplot2 plot must begin with:

```{r, eval = F}
ggplot()
```

## Basics

Geoms and aesthetics are the two most important things to understand.

Geoms are the elements we want to plot (lines, points, bars, etc).

Aesthetics are how we'd like geoms to appear (where, what colour, what size, etc).

## Geoms

**Geoms** are visual elements that will represent variables in a data frame. For example, points, bars, and lines.

Geoms are created with functions of the form `geom_*()`. For example, `geom_point()` will add points.

We add geoms in layers to our plot using `+`. E.g.

```{r, eval = F}
ggplot() + geom_point()
```

And we can add multiple geoms:

```{r, eval = F}
ggplot() + geom_point() + geom_line() ...
```

## Adding a data frame

ggplot2 references data frames using the `data` argument.

This can be used within `ggplot()` or `geom_*()`

Examples:

```{r, eval = F}
ggplot(data = mtcars) + geom_point()

ggplot() + geom_point(data = mtcars)
```

## Aesthetics

**Aesthetics** map variables in the data frame to aspects of the geoms. Typical geom aspects that we map variables to include the x and y coordinates, colour, and shape.

Aesthetics are defined within `ggplot()` or geom functions via the function `aes()` and we reference variable names **without** quotation marks.

E.g., `ggplot(my_data, aes(x = my_var))` will map values of the variable `my_var` (from the `my_data` data frame) to the x coordinates of geoms.

## Aesthetics

Different aesthetics can be used for different geoms. Common ones include `x`, `y`, `color`, `shape`.

To find out which aesthics can be used for a certain geom, read the "Aesthetics" section on the help page for that geom.

For example, look at the help page for `geom_point()`.

## Aesthetics

For example, this is a plot with points (geoms) whose x coordinates are determined by the variable `hp`, and whose `y` coordinates are defined by the variable `mpg`, which come from the `mtcars` data frame:

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point()
```

## Aesthetics

```{r}
ggplot() + geom_point(data = mtcars, aes(x = hp, y = mpg))
```

## Test yourself

Using the `mtcars` data, create a scatter plot of cars' weight (`wt`) and horsepower (`hp`).

Create the same plot by using the `data` argument and aesthetic function (`aes()`) in `ggplot()` or `geom_point()`

## Scalar aesthetics 

We can also use scalar values for aesthetics. In such cases, we must make the changes within `geom_*()` and **outside** of `aes()`.

For example, we can map a variable to the `color` aesthetic:

```{r}
ggplot(mtcars, aes(x = hp, y = mpg, color = hp)) +
  geom_point()
```

## Scalar aesthetics 

However, we can also make all the points a single colour:

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "red")
```

## Test yourself

Create another scatter plot of horsepower (`hp`) and Miles per Gallon (`mpg`).

Adjust the `size` aesthetic of the points in two ways:

1. Adjust the size depending on the value of `mpg`.
2. Make all points the same size of 4.

## Using ggplot() or geom_*()?

When a data frame or aesthetic is placed in `ggplot()`, it is used by all geom layers that don't override those values.

```{r}
ggplot(mtcars, aes(x = hp, y = mpg, color = hp)) +
  geom_point() +
  geom_line()
```

## Using ggplot() or geom_*()?

Data or aesthetics used within `geom_*()` will be used by that geom only.

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(aes(color = hp)) +
  geom_line()
```

## Test Yourself

Using the scatter and line plot of `hp` and `mpg`, use `mpg` to adjust the following aesthetics:

- `color` of both geoms.
- `alpha` of the line.
- `size` of the points.

## Test Yourself

Answer:

```{r}
ggplot(mtcars, aes(x = hp, y = mpg, color = mpg)) +
  geom_point(aes(size = mpg)) +
  geom_line(aes(alpha = mpg))
```

## Thinking in terms of the data

What if we wanted to visualise the means of groups like in this example?

```{r, echo = F, message = F, fig.width=5, fig.height=4}
library(dplyr)
mtcars %>% 
  group_by(cyl) %>% 
  summarise(mpg = mean(mpg)) %>% 
  ggplot(aes(x = cyl, y = mpg)) +
    geom_bar(stat = "identity")
```

## Thinking in terms of the data

In terms of data, we're using `mtcars`, so `ggplot(data = mtcars)`

In terms of geoms, we're seeing bars, so that's `geom_bar()`.

In terms of aesthetics:

- x-axis defined by `cyl`, so `aes(x = cyl)`.
- y-axis defined by `mpg`, so `aes(x = cyl, y = mpg)`

Right??

## Thinking in terms of the data

Wrong :(

The bar heights are **means**, not single observations.

So we need a data frame that contains this information!

## Thinking in terms of the data

```{r, message = F}
library(dplyr)
d <- mtcars %>% 
  group_by(cyl) %>% 
  summarise(mpg = mean(mpg))
d
```

## Thinking in terms of the data

```{r, message = F, fig.width=5, fig.height=4}
ggplot(d, aes(x = cyl, y = mpg)) +
  geom_bar(stat = "identity")
```

## Thinking in terms of the data

```{r, message = F, fig.width=5, fig.height=4}
ggplot(d, aes(x = cyl, y = mpg, label = round(mpg, 2))) +
  geom_bar(stat = "identity") +
  geom_text(vjust = 1.5, color = "white")
```

# Challenge

## Challenge

Create the following plot using the `mtcars` data:

```{r, echo = F, message=F, fig.width=5, fig.height=5}
library(ggplot2)
library(dplyr)
id <- mtcars
gd <- id %>%
  group_by(cyl) %>% 
  summarise(hp = mean(hp))

ggplot(id, aes(x = factor(cyl), y = hp, color = cyl, fill = cyl)) +
  geom_point() +
  geom_bar(data = gd, stat = "identity", alpha = .2) +
  guides(color = "none", fill = "none") +
  labs(
    x = "Number of cylinders",
    y = "Horsepower"
  ) +
  theme_bw()
```
