---
title: 'Using R: ggplot2'
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

## What/Why ggplot2?

`ggplot2` is a package for **plotting**.

Many would consider it to be R's most powerful data vizualisation tool.

It's incredibly flexible, giving you control over almost every aspect of a plot.

But with flexibility comes the challenge of mastery.

## Some examples

```{r,echo = F, message = F}
library(dplyr)
library(ggplot2)

# Car horsepower
id <- mtcars %>%
  tibble::rownames_to_column() %>% 
  mutate(cyl = factor(cyl, levels = c(4, 6, 8)))

gd <- id %>% 
        group_by(cyl) %>% 
        summarise(
          hp = mean(hp)
        )

ggplot(id, aes(x = cyl, y = hp, color = cyl, fill = cyl)) +
  geom_bar(data = gd, stat = "identity", alpha = .3) +
  ggrepel::geom_text_repel(aes(label = rowname), color = "black", size = 2.5, segment.color = "grey") +
  geom_point() +
  guides(color = "none", fill = "none") +
  theme_bw() +
  labs(
    title = "Car horespower by cylinders",
    x = "Number of cylinders",
    y = "Horsepower"
  )
```

## Some Examples

```{r, echo = F, message = F}
# Step 1: Fit the data
d <- mtcars
fit <- glm(vs ~ hp, family = binomial(), data = d)

# Step 2: Obtain predicted and residuals
d$predicted <- predict(fit, type="response")
d$residuals <- residuals(fit, type = "response")

# Steps 3 and 4: plot the results
ggplot(d, aes(x = hp, y = vs)) +
  geom_segment(aes(xend = hp, yend = predicted), alpha = .2) +
  geom_point(aes(color = residuals), size = 3) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red") +
  guides(color = FALSE) +
  geom_point(aes(y = predicted), shape = 1, size = 3) +
  theme_bw() +
  labs(
    x = "Horespower",
    y = "V engine (or S)",
    title = "Predicted and residual values from logistic regression"
    
  )
```

## Some examples

```{r, echo = F, message = F}
iris %>%
  select(Species, Petal.Length, Petal.Width) %>% 
  rename(Length = Petal.Length, Width = Petal.Width) %>% 
  mutate(id = row_number()) %>% 
  tidyr::gather(var, val, -id, -Species) %>% 
  ggplot(aes(x = var, y = val, group = id, color = Species)) +
  geom_line(size = 1.5, alpha = .1) +
  geom_point(size = 4, alpha = .2) +
  scale_color_brewer(palette = "Set1") +
  theme_bw() +
  labs(
    x = "Petal Size",
    y = "cm",
    title = "Petal sizes of iris flowers"
  )
```

## Some examples

```{r, echo = F, message = F}
d <- iris[-5]

# Univariate
ggplot(iris, aes(x = Sepal.Width, fill = Species)) +
  geom_histogram(data = d, fill = "grey", alpha = .5) +
  geom_histogram(colour = "black") +
  facet_wrap(~ Species) +
  guides(fill = FALSE) +
  theme_bw() +
  labs(
    x = "Sepal Width",
    y = NULL,
    title = "Distribution of Sepal Width for different iris species")
```

## Some examples

```{r, echo = F}
airquality %>% 
  select(Ozone, Temp) %>% 
  na.omit() %>% 
  ggplot(aes(x = Ozone, y = Temp, color = Temp)) +
  geom_point(size = 4) +
  guides(color = "none") +
  scale_colour_gradientn(colours = c("deepskyblue2", "chocolate1", "firebrick")) +
  theme_classic() +
  theme(panel.background = element_rect(fill = "black")) +
  labs(
    x = "Mean ozone in parts per billion",
    y = "Maximum daily Temperature (degrees F)",
    title = "Ozone levels correspond to temperature"
  )

```

## Some examples

```{r, echo = F, message = F}
library(ourworldindata)

d <- financing_healthcare %>% 
  filter(year == 2013) %>% 
  select(country, health_exp_total, gdp) %>% 
  na.omit() %>% 
  mutate(exp_gdp = 100 * health_exp_total / gdp) %>% 
  rename(region = country) %>% 
  mutate(region = ifelse(region == "United States", "USA", region))

world_map <-  map_data(map = "world") %>% left_join(d)

ggplot(world_map) +
  geom_map(map = world_map, aes(map_id = region, x = long, y = lat, fill = exp_gdp)) +
  scale_fill_gradient(name = "Healthcare expenditure\nas % of GDP",
                      low = "palegreen1", high = "navyblue", na.value = "gray95") +
  coord_equal() +
  ggtitle("Total Healthcare Expenditure as Share (%) of National GDP, 2013") +
  theme_void()
```

## Thinking in ggplot2

Using `ggplot2` often requires a mental shift when it comes to plotting.

When we use tools like SPSS or Excel, we often try various plotting methods until something gives us what we want. But we seldom play with the data.

When it comes to ggplot2, we're forced (in a good way) to manipulate the data. This can take some time to get used to.

## Thinking in ggplot2

The logic behind ggplot2 is to map values from variables (columns) in a data frame to aesthetics of visual elements (geoms) in a plot.

This will make more sense as we progress.

## Basics

(Install and) Load the `ggplot2` package:

```{r}
# install.packages("ggplot2")
library(ggplot2)
```

A ggplot2 plot must begin with:

```{r, eval = F}
ggplot()
```
