---
title: 'Using R: ggplot2 themes, labels, and other features'
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 3,
  fig.align = "center"
)
```

## Beyond geoms and aesthetics

Geoms and aesthetics are how we represent data in the plot.

However, there are many other aspects of our plots we might want to make adjustments too.

For example, the background colour, the axis ticks and labels.

# Themes

## Themes

The visual elements outside of geoms are mostly controlled by themes.

Let's create a default plot to work with:

```{r}
library(ggplot2)
p <- ggplot(mtcars, aes(x = hp, y = mpg)) + geom_point()
p
```

## Themes

There are many existing themes we can apply. For example:

```{r}
p + theme_bw()
```

## Themes

```{r}
p + theme_classic()
```

## Themes

See the complete list of themes via the help page for any theme.

e.g., `?theme_bw`

## Test yourself

Create the plot using a theme we haven't used yet.

## Custom themes

We can also customise the theme elements using `theme()`.

See `?theme` for a complete list of elements that can be adjusted.

## Custom themes

```{r}
p + theme(panel.background = element_rect(fill = "skyblue"))
```

## Custom themes

```{r}
p + theme(axis.line.x = element_line(color = "blue",
                                     size = 3,
                                     lineend = "round"),
          axis.line.y = element_line(color = "orange",
                                     size = 2,
                                     lineend = "square")
          )
```

# Labels

## Labels

We almost always want to change the plot title and axis labels.

The best way to do this is using `+ labs()`, and providing three arguments:

- `title = ""`
- `x = ""`
- `y = ""`

## Labels

```{r}
p + labs(x = "Horsepower", y = "Miles Per Gallon")
```

## Test yourself

Add a title to the previous plot.