---
title: 'Using R: understanding ggplot2 layers'
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
knitr::opts_knit$set(root.dir = '../../../../')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 3,
  fig.width = 3,
  fig.align = "center"
)
```

## Layers

When we add (`+`) geoms to a plot, this is done in layers.

geoms added earlier in the plot chain will appear **behind** geoms added later.

Imagine drawing each geom on a plastic sheet. When you add it to the plot, you put it down on top of the previous one.

## Examples

Here the points are added ontop of the line:

```{r}
library(ggplot2)
ggplot(mtcars, aes(x = hp, y = wt)) +
  geom_line(size = 2) + 
  geom_point(size = 2, color = "red")
```

## Examples

Here, the line is added ontop of the points:

```{r}
ggplot(mtcars, aes(x = hp, y = wt)) +
  geom_point(size = 2, color = "red") +
  geom_line(size = 2)
```

## Some uses

Understanding layers is important from the point of view that you will often make simple errors.

But carefully layering our geoms can also prove useful for many reasons.

## Examples

Let's examine ozone levels in May using the `airquality` data set.

```{r, message=F}
library(dplyr)
d <- airquality %>% filter(Month == 5)
head(d)
```


## Example

Let's draw points:

```{r, warning=F, fig.width=8}
ggplot(d, aes(x = Day, y = Ozone)) +
  geom_point()
```

## Example

How might we make these large white points with black outlines?

## Example

```{r, fig.width=8, warning=F}
ggplot(d, aes(x = Day, y = Ozone)) +
  geom_point(size = 3) +
  geom_point(color = "white", size = 2)
```

## Example

```{r, fig.width=8, warning=F}
ggplot(d, aes(x = Day, y = Ozone)) +
  geom_point(size = 10, color = "blue") +
  geom_point(size = 8,  color = "green") +
  geom_point(size = 6,  color = "yellow") +
  geom_point(size = 4,  color = "orange") +
  geom_point(size = 2,  color = "red")
```

## Test Yourself

Add a line to one of the previous Ozone plots.

Make sure that the dots appear on top of the line.

Bonus: change the `linetype` aesthetic on the line to equal 2.

## Test Yourself

Answer

```{r, fig.width=8, warning=F}
ggplot(d, aes(x = Day, y = Ozone)) +
  geom_line(linetype = 2) +
  geom_point(size = 3) +
  geom_point(color = "white", size = 2)
```

# Challenge

## Challenge

In this challenge we'll use the tracking data:

```{r}
d <- read.csv("data/tracking.csv")
```

This is a long-format data file containing people's (`id`) tracking time (`time`) on four trials (`trial`). On all four trials, participants track EITHER a Box or a Circle (`Shape`).

## Challenge

Plot everyone's tracking times over the four trials via:

```{r, fig.width=8, fig.height=3}
ggplot(d, aes(x = trial, y = time, color = Shape)) +
  geom_line(aes(group = id), alpha = .2)
```

## Challenge

Add group means so the plot looks similar to:

```{r, echo = F, fig.width=8, fig.height=5}
gd <- d %>% 
  group_by(Shape, trial) %>% 
  summarise(time = mean(time))

ggplot(d, aes(x = trial, y = time, color = Shape)) +
  geom_line(aes(group = id), alpha = .2) +
  geom_line(data = gd, aes(group = Shape), size = 2) +
  geom_point(data = gd, size = 6) +
  geom_point(data = gd, size = 4, color = "white") +
  theme_bw()
```

## Challenge

Answer

```{r, eval = F}
gd <- d %>% 
  group_by(Shape, trial) %>% 
  summarise(time = mean(time))

ggplot(d, aes(x = trial, y = time, color = Shape)) +
  geom_line(aes(group = id), alpha = .2) +
  geom_line(data = gd, aes(group = Shape), size = 2) +
  geom_point(data = gd, size = 6) +
  geom_point(data = gd, size = 4, color = "white") +
  theme_bw()
```