---
title: "Functional Science"
author: "Dr Simon Jackson"
date: "Knit on `r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation
---

```{r setup, echo = FALSE, warning = FALSE}
require(knitr, quietly = TRUE)
opts_knit$set(root.dir = '../../')

library(png)
library(grid)
```

## Dr Simon Jackson

```{r, echo = FALSE, warning = FALSE, fig.width = .5, fig.height = .5}
img <- readPNG("Rmd/img/social_media_icons/gmail.png")
grid.raster(img)
```
drsimonjackson@gmail.com

```{r, echo = FALSE, warning = FALSE, fig.width = .5, fig.height = .5}
img <- readPNG("Rmd/img/social_media_icons/twitter.png")
grid.raster(img)
```
[\@drsimonj](https://twitter.com/drsimonj)

```{r, echo = FALSE, warning = FALSE, fig.width = .5, fig.height = .5}
img <- readPNG("Rmd/img/social_media_icons/github.png")
grid.raster(img)
```
[github.com/drsimonj](https://github.com/drsimonj)

<br><br>

**Slides available at my Github**  
Follow `functional-science / Rmd / slides`

# What is Functional Science?

## Functional Programming

The idea comes from thinking like a programmer, and a *functional programming* philosophy.

In computer science, functional programming is a programming paradigm — a style of building the structure and elements of computer programs — that treats computation as the evaluation of mathematical functions and avoids changing-state and mutable data.

In functional code, the output value of a function depends only on the arguments that are input to the function, **so calling a function *f* twice with the same value for an argument *x* will produce the same result *f(x)* each time**.

## Functional Science?

A style of conducting science that is:

- Reliable/replicable
- Valid
- Sharable
- Generalisable
- Efficient

# The Golden Rules

## Rule 1: Outsource your memory

Never depend on your memory to answer questions like:

- What is this variable or how did I calculate it?
- Why did I make this hypothesis or run this analysis?
- Which file was I working on last or had the data I need?

## Rule 2: Automate as much as you can

Avoid having to do the same task over and over again. E.g.:

- Write the same question stem.
- Parse data
- Copy and paste results and then format the table.

## Rule 3: Be simple and be consistent

Decide on an intuitive system and stick with it!

- Folder structures
- Naming files
- Naming variables
